HO unification from object language to meta language

### 2024-02-26 paper/main.tex
update tex
wip

Our encoding of DTT may look ``semi shallow'' since we we use the meta language
lambda abstraction but not its application (for the terms of type \elpiIn{tm}).
A fully shallow encoding unfortunately does not fit our use case, although
it would make the running example work:

@@

\begin{figure}[H]
  \begin{elpicode}
    kind tm type.
    type app list tm -> tm.
    type lam (tm -> tm) -> tm.
    type con string -> tm.
  \end{elpicode}
  \caption{Common terms}
  \label{code:common-terms}
  \Description[code:common-terms]{code:common-terms}
\end{figure}
%
\noindent
This encoding is very similar to the one introduced in \cref{sec:intro},\todo{say we do not care about types} except
for the \elpiIn{all} constructor. We explicitly neglect it since blabla.
Moreover, since we are working with unification variables we need to introduce a
new constructor.

In the case of the OL, variables have no scope. For example, the subterm
\coqIn{P x} from the instance forall\_dec has \coqIn{P} an higher order
variables with type \coqIn{A ~$\to$~ Prop} and \coqIn{x} as bound variable. 
However, at the meta level, the translation of \coqIn{P x} becomes
\elpiIn{app[P, x]}, that is, \elpiIn{P} cannot reference \elpiIn{x}. In our
encoding, we represent the variables of the OL in the following way:

\begin{elpicode}
  type fo_uv nat -> tm.
\end{elpicode}

\noindent 
In particular, a variable of the OL is identified by a integer
refereeing a memory address. This constructor is prefixed with \elpiIn{fo} since
the variable, having no scope, is a first-order variable. Finally, the memory of
the OL is depicted by the following type abbreviation:

\begin{elpicode}
  typeabbrev fo_subst list (option tm).
\end{elpicode}

@@

On the other hand, the terms of the ML have exactly the same shape
has the terms of \cref{code:common-terms}. However, since the ML is an Higher
Order Programming Language, unification unification variables must have a scope.
We represent them in the following way:

\begin{elpicode}
  type ho_uv nat -> list tm -> tm.
\end{elpicode}

Of course some attention
should be payed when dealing with \elpiIn{ho_uv}, since we have to certify
that an \elpiIn{uv i} remain in the pattern fragment, that is, the
list of term in the scope of \elpiIn{i} is always a list of distinct names.

Another important implementation should be provided in order to represent the
lambda abstractions and the applications of the ML. As outlined in
\cref{sec:encodings}, <<\elpiIn{x\ f x}>> has not the same semantic as 
<<\elpiIn{lam x\ app[f, x]}>>, since, even though, the two terms represent the same
concept, the former is the lambda abstraction of \elpiIn{f} applied to the 
binder in ML, whereas the latter is the same term representation but wrt the OL.
We already have defined the nodes \elpiIn{app} and \elpiIn{lam} for the OL,
therefore, in the following code snippet, we give their representation in the
ML.

\begin{elpicode}
  kind assmt type.
  type abs (tm -> assmt) -> assmt.
  type val tm -> assmt.
\end{elpicode}

\noindent
In particular, the node \elpiIn{abs} stands for the lambda abstraction. The
node \elpiIn{val} contains terms of the object language. The node for variable
application is not really necessary, since, using the syntax of the ML, it is
defined by putting two terms of the ML side by side. For example, the terms 
<<\elpiIn{x\ f x}>> becomes <<\elpiIn{abs x\ f x}>>.

The memory of the ML, is defined with the following type abbreviation:

\begin{elpicode}
  typeabbrev ho_subst list (option tm).
\end{elpicode}

@@

\begin{elpicode}
  type fo_deref fo_subst -> tm -> tm -> o.
  fo_deref S (app L1) (app L2) :- forall2 (fo_deref S) L1 L2.
  fo_deref S (lam F1) (lam F2) :- 
    pi x\ fo_deref S x x => fo_deref S (F1 x) (F2 x).
  fo_deref _ (con X) (con X).
\end{elpicode}

@@

\begin{elpicode}
  type ho_deref ho_subst -> tm -> tm -> o.
\end{elpicode}

\begin{elpicode}
  fo_deref S (fov N) R :- assigned? N S T, fo_deref S T R.
  fo_deref S (app [fov N|L]) R :- 
    assigned? N S T, beta T L R', fo_deref S R' R.
  fo_deref _ (fov N) (fov N).
\end{elpicode}

@@

@@

@@

% \subsection{implementation}

% l'HO encoding e' esattamente lambda Prolog/elpi, il compilatore
% in pratica potrebbe essere scritto in un meta language, qui lo si presenta in
% elpi stesso. HO e FO in questo paper sono deep embedded in elpi per parlarne,
% ma in pratica il nostro solever, prendere XX, scrivere la clausola compilata.

% \section{recovering eta}

% \begin{elpicode}
% q (all x\ F x) = q (all x\ app[f,x]) /\ p f = p F
% F = fun a => app [f,a] ----> F = f
% \end{elpicode}

% l'utene da p su f, mentre l'istanza pe q forza F a fun ..

% \section{recovering beta}

% \begin{elpicode}
%   q (all x\ F x) = q (all x\ app[f,x,x]) /\ p1 (app[f,a,a]) = p1 (app[F,a])
%   F = fun y => app [f,y,y] ----> (app[F,a]) ~> app[f, a, a].
% \end{elpicode}

% qui la sintesi di F puo generare un beta redex, quindi ci mettiamo
% p1 F1, e decomp beta F [a] F1.

% \section{recovering eta-beta within unification (non linear variables)}

% se i problemi di cui sopra avvengono nello stesso termine

% \begin{elpicode}
%   q2 (all x\ F x) (app[F,a]) = q2 (all x\ app[f,x,x]) (app[f,a,a])
% \end{elpicode}

% bisogna slegare le due F e poi unificare le soluzioni tra di loro

% \section{heuristic / binary app}

% fo approx / sub pattern fragment

% \begin{elpicode}
%   p (all x\app[F,x,a]) (app[F,b]) = p (all x\app[f,x,x,a]) (app[f,b,b])
%   p (all x\G x) F' = 
%   G = x\ f x x a
%   F = lam x\f x x
%   F' = (app[f,b,b])
%   link (F a) F'
%   link G F
% \end{elpicode}

% % questo esempio recupera l'app binaria perche

% \begin{elpicode}
%   (app (app F x) a) = (app (f x x) a)
% \end{elpicode}

% \section{Stats}

% Concretamente:

% - quante volte l'ordine superiore è necessario in stdpp/tlc

% \section{Conclusion}

>>>>

Paper \cite{10.1145/2966268.2966272} introduces semi-shallow.

Our encoding of DTT may look ``semi shallow'' since we use the meta-language
lambda abstraction but not its application (for the terms of type \elpiIn{tm}).
A fully shallow encoding unfortunately does not fit our use case, although
it would make the running example work:

@@

\begin{figure}[H]
  \begin{tabular}{ll}
  \begin{minipage}{0.21\textwidth}
  \begin{elpicodetab}
kind fm type.
type fapp list fm -> fm.
type flam (fm -> fm) -> fm.
type fcon string -> fm.
type fuva nat -> fm.
\end{elpicodetab}
  \end{minipage}
  &
  \begin{minipage}{0.24\textwidth}
  \begin{elpicodetab}
kind tm type.
type app list tm -> tm.
type lam (tm -> tm) -> tm.
type con string -> tm.
type uva nat -> list tm -> tm.
  \end{elpicodetab}
  \end{minipage}
  \end{tabular}
  \caption{Source and Target language}
  \label{code:common-terms}
  \Description[code:common-terms]{code:common-terms}
\end{figure}

\noindent
This encoding is very similar to the one introduced in \cref{sec:intro},\todo{say we do not care about types} except
for the \elpiIn{all} constructor. We explicitly neglect it since blabla.
Moreover, since we are working with unification variables we need to introduce a
new constructor.

In the case of the OL unification variables \elpiIn{fuv} have no scope.
For example the statement of the instance \coqIn{forall_dec} contains the
term \coqIn{P x} where \coqIn{P} stands for a function of type
\coqIn{A ~$\to$~ Prop}. However our representation of \coqIn{P x} in the meta
language is \elpiIn{app[P, x]} where \elpiIn{P} is of type \elpiIn{ftm}.
In the terms of the ML unification variables come with a scope.

\begin{elpicode}
  type uva nat -> list tm -> tm.
\end{elpicode}

\noindent
Of course some attention
should be payed when dealing with \elpiIn{uva}, since we have to certify
that an \elpiIn{uva i} remain in the pattern fragment, that is, the
list of term in the scope of \elpiIn{i} is always a list of distinct names.

Unification variables of the OL are identified by a natural number
referring to memory address. 
Finally, the memory of
the OL is depicted by the following type abbreviation:

\begin{elpicode}
  typeabbrev fsubst list (option fm).
\end{elpicode}

@@

Another important implementation should be provided in order to represent the
lambda abstractions and the applications of the ML. As outlined in
\cref{sec:encodings}, <<\elpiIn{x\ f x}>> has not equal to 
<<\elpiIn{lam x\ app[f, x]}>>, since, even though the two terms represent the same
concept, the latter uses the abstraction and application of the OL.

\begin{elpicode}
  kind assmt type.
  type abs (tm -> assmt) -> assmt.
  type val tm -> assmt.
  typeabbrev subst list (option tm).
\end{elpicode}

\noindent
In particular, the node \elpiIn{abs} stands for the lambda abstraction of the ML.
The
node \elpiIn{val} contains terms of the object language. The node for variable
application is not really necessary, since, using the syntax of the ML, it is
defined by putting two terms of the ML side by side. For example, the terms 
<<\elpiIn{x\ f x}>> becomes <<\elpiIn{abs x\ f x}>>.

The memory of the ML is represented by a \elpiIn{subst}.

@@

\begin{elpicodetab}
  type fderef fsubst -> fm -> fm -> o.
  fderef S (fuva N) R :- assigned? N S T, fderef S T R.
  fderef S (app [fuva N|L]) R :- 
    assigned? N S (app NL), append NL L AL, fderef S (app AL) R.
  fderef _ (fuva N) (fuva N).
  fderef S (fapp L1) (fapp L2) :- forall2 (fderef S) L1 L2.
  fderef S (flam F1) (flam F2) :- 
    pi x\ fderef S x x => fderef S (F1 x) (F2 x).
  fderef _ (fcon X) (fcon X).
\end{elpicodetab}

@@

\begin{elpicode}
  type deref subst -> tm -> tm -> o.
  deref S (uva N L) R :- assigned? N S T, beta-abs T L T', deref S T' R.
\end{elpicode}

@@

\noindent
\otext{An other example: \\
  \elpiIn{@lam x\ @app[@f, @app[@X, x]] = @Y, (@lam x\ x) = @X.}}

@@

\subsection{Tricky examples}

\begin{elpicode}
  triple ok (@lam x\ @app[@f, @app[@X, x]]) @Y,
  triple ok @X (@lam x\ x),
  triple ok @Y @f
\end{elpicode}

@@
